---
title: "Sleep Health & Lifestyle – Exploratory Analysis"
author: "Pascal Hermann Kouogang Tafo"
format: html
editor: visual
---

## INTRODUCTION

This assignment analyses the **Sleep Health & Lifestyle Dataset** to understand how lifestyle factors such as occupational role and psychological stress could shape two core sleep outcomes: sleep duration and self-rated sleep quality.Understanding these relationships is relevant for us as individual and corporations when creating workplace wellness programs to improve employees performance.

## APPROACH

To conduct the analysis of the dataset, i will implement a structured data science pipeline using the "tidyverse" framework as followed:

1)  Load the CSV dataset in R and commit to an existing GitHub repository ensuring its accessibility at anytime .

2)  Rename some variables by removing "\_" and follow a consistent naming convention.

3)  Transform my dataset from Wide to Long format using the **pivot_longer** function to convert the sleep duration and quality into a single metric column for easier modelling and faceted plotting.

4)  Analyze and Visualize the Correlation between Sleep Duration and Quality related to Occupation. Interpret the result

5)  Analyze and Visualize the Correlation between Sleep Duration and Quality related to Stress Level. Interpret the result

## Load Library

```{r Load-R-packages}


library(tidyverse)
library(janitor) # for cleaning column names

```

## Load and clean the data

I will read the file from my GitHub. For the cleaning i will need the help of the LLM Claude because some columns names are inconsistent and some variables are compounds. Here is the prompt : **"Clean this dataset by renaming to snake_case and parsing compounds variables."**

```{r}

# Read file

url <- "https://raw.githubusercontent.com/Pascaltafo2025/PROJECT-2--TIDY-DATA-ANALYSIS/refs/heads/main/Sleep_health_and_lifestyle_dataset.csv"

Sleep_health_and_lifestyle <- read_csv(url)

head(Sleep_health_and_lifestyle,10)

## Rename to snake_case to insure consistency

Sleep_health_and_lifestyle <- Sleep_health_and_lifestyle |>
  rename(
    person_id             = `Person ID`,
    gender                = Gender,
    age                   = Age,
    occupation            = Occupation,
    sleep_duration_hrs    = `Sleep Duration`,
    sleep_quality_score   = `Quality of Sleep`,
    physical_activity_min = `Physical Activity Level`,
    stress_level          = `Stress Level`,
    bmi_category          = `BMI Category`,
    blood_pressure        = `Blood Pressure`,
    heart_rate_bpm        = `Heart Rate`,
    daily_steps           = `Daily Steps`,
    sleep_disorder        = `Sleep Disorder`
  )

# Clean 
Sleep_health_and_lifestyle_clean <- Sleep_health_and_lifestyle |>
  mutate(
    # Decision 1: harmonise synonymous BMI labels
    
    bmi_category  = recode(bmi_category, "Normal" = "Normal Weight"),
    bmi_category  = factor(bmi_category,
                           levels = c("Normal Weight", "Overweight", "Obese")),
    # Decision 2: NA sleep_disorder → explicit "None"
    sleep_disorder = replace_na(sleep_disorder, "None"),
    sleep_disorder = factor(sleep_disorder,
                            levels = c("None", "Insomnia", "Sleep Apnea")),
    # Decision 3: parse compound blood pressure string
    bp_systolic   = as.integer(str_split_fixed(blood_pressure, "/", 2)[, 1]),
    bp_diastolic  = as.integer(str_split_fixed(blood_pressure, "/", 2)[, 2]),
    # Decision 4: type coercions
    gender        = factor(gender),
    occupation    = factor(occupation),
    stress_level  = as.integer(stress_level),
    person_id     = as.character(person_id)
  ) |>
  select(-blood_pressure)

head(Sleep_health_and_lifestyle_clean,10)

```

# Transform my dataset from Wide to Long format using the **pivot_longer** function

```{r}

Sleep_health_and_lifestyle_clean_long <- Sleep_health_and_lifestyle_clean |>
  pivot_longer(
    cols      = c(sleep_duration_hrs, sleep_quality_score),
    names_to  = "sleep_metric",
    values_to = "metric_value"
  ) |>
  mutate(
    sleep_metric = recode(sleep_metric,
      "sleep_duration_hrs"  = "Sleep Duration (hrs)",
      "sleep_quality_score" = "Sleep Quality (score)"
    )
  )

head(Sleep_health_and_lifestyle_clean_long,10)

```

## Analyze and Visualize the Correlation between Sleep Duration and Quality related to Occupation

```{r}

# Compute sleep statistical summary by occupation

occupation_summary <- Sleep_health_and_lifestyle_clean %>%
  group_by(occupation) %>%
  summarise(
    n                   = n(),
    mean_sleep_duration = mean(sleep_duration_hrs,  na.rm = TRUE),
    mean_sleep_quality  = mean(sleep_quality_score, na.rm = TRUE),
    mean_stress         = mean(stress_level,         na.rm = TRUE),
    .groups = "drop"
  )
occupation_summary


```

## Stress-level averages

```{r}
stress_summary <- Sleep_health_and_lifestyle_clean %>%
  group_by(stress_level) %>%
  summarise(
    n                   = n(),
    mean_sleep_duration = mean(sleep_duration_hrs,  na.rm = TRUE),
    mean_sleep_quality  = mean(sleep_quality_score, na.rm = TRUE),
    .groups = "drop"
  )

stress_summary
```

## Visualization

```{r}
# Overall Pearson correlation
overall_r <- cor.test(Sleep_health_and_lifestyle_clean$sleep_duration_hrs, Sleep_health_and_lifestyle_clean$sleep_quality_score)
cat(sprintf("Overall Pearson r = %.3f  (p = %.2e)\n\n",
            overall_r$estimate, overall_r$p.value))


# =============================================================================
# STEP 7 – Visualisation: Sleep Metrics ~ Occupation
# =============================================================================

# ── OCCUPATION PLOT A: raw scatter coloured by occupation ────────────────────
occ_palette <- c(
  "Accountant"          = "#4C72B0",
  "Doctor"              = "#DD8452",
  "Engineer"            = "#55A868",
  "Lawyer"              = "#C44E52",
  "Manager"             = "#8172B2",
  "Nurse"               = "#937860",
  "Sales Representative"= "#DA8BC3",
  "Salesperson"         = "#E7BA52",
  "Scientist"           = "#8C8C8C",
  "Software Engineer"   = "#CCB974",
  "Teacher"             = "#64B5CD"
)

ggplot(Sleep_health_and_lifestyle_clean,
  aes(x = sleep_duration_hrs,
      y = sleep_quality_score,
      colour = occupation)) +
  geom_point(alpha = 0.55, size = 2.2) +
  geom_smooth(data   = Sleep_health_and_lifestyle_clean,
              aes(x  = sleep_duration_hrs,
                  y  = sleep_quality_score),
              method = "lm", formula = y ~ x,
              colour = "black", linewidth = 0.9,
              linetype = "dashed", se = TRUE, inherit.aes = FALSE) +
  scale_colour_manual(values = occ_palette, name = "Occupation") +
  scale_x_continuous(breaks = seq(5, 9, 0.5)) +
  scale_y_continuous(breaks = 1:10) +
  labs(
    title    = "Sleep Duration vs Quality of Sleep by Occupation",
    subtitle = sprintf("Individual observations  |  Overall Pearson r = %.2f",
                       overall_r$estimate),
    x        = "Sleep Duration (hours per night)",
    y        = "Quality of Sleep (1 = Poor -> 10 = Excellent)",
    caption  = "Source: Sleep Health & Lifestyle Dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    plot.subtitle   = element_text(colour = "grey40"),
    legend.position = "right"
  )

# ── OCCUPATION PLOT B: grouped bar chart – mean duration & quality ────────────
occ_long <- occupation_summary %>%
  pivot_longer(cols = c(mean_sleep_duration, mean_sleep_quality),
               names_to  = "metric",
               values_to = "mean_value") %>%
  mutate(
    metric = recode(metric,
      "mean_sleep_duration" = "Avg Sleep Duration (hrs)",
      "mean_sleep_quality"  = "Avg Sleep Quality (score)"
    ),
    occupation = fct_reorder(occupation, mean_value, .fun = max)
  )
 ggplot(occ_long,
  aes(x    = mean_value,
      y    = occupation,
      fill = metric)) +
  geom_col(position = "dodge", width = 0.65, alpha = 0.88) +
  scale_fill_manual(values = c("Avg Sleep Duration (hrs)"  = "#4C72B0",
                               "Avg Sleep Quality (score)" = "#DD8452"),
                    name   = "Sleep Metric") +
  scale_x_continuous(limits = c(0, 10), breaks = seq(0, 10, 2)) +
  labs(
    title   = "Mean Sleep Duration & Quality by Occupation",
    x       = "Mean Value",
    y       = NULL,
    caption = "Source: Sleep Health & Lifestyle Dataset"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )


```
